---
title: "Introduction to R"
author: "Mairead Shaw"
editor: visual
format: pdf
---

## Introduction to R

This document is a pared-down version of the `part1.R` and `part2.R` files, intended for a shorter workshop. There is no new content, just a light reorganization to focus less on basic data structures and more on getting up and running using packages and conducting analyses, intended to be delivered as a single 90-minute workshop.

## Additional Resources

These materials serve as a quick introduction. There are many excellent resources for learning R online. Two I will draw your attention to are (1) [R for Data Science](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund, an excellent resource to learn a great deal of R, and (2) the [R Graph Gallery](https://r-graph-gallery.com/) curated by Yan Holtz, which shows examples of all kinds of graphs and how to code them.

## Agenda

-   Why should we code?
-   Creating projects
-   Variables, vectors, indexing vectors
-   Loading files -- absolute vs relative filepaths
-   Installing and loading packages -- `dplyr`, `sjPlot`
-   Inspecting data: `head()`, `tail()`, `str()`, `View()`
-   Functions
-   `dplyr` verbs: `filter()`, `select()`, `mutate()`, `group_by()`, `arrange()`
-   Statistical tests: t-test, ANOVA, post-hoc testing, linear and logistic regression
-   If time, graphing with `ggplot2`

## Why Should We Code?

We code to accomplish tasks that would be less efficient or impossible to do by hand. There are many different coding languages. Which one you use depends on the task at hand and the conventions of a given field. A lot of programming languages are flexible and *can* do a lot, but have different things that they excel at. I like to think of coding languages as being like utensils: they help us accomplish a task, and there are a lot of options, but some utensils are better at some jobs (e.g., spoon vs fork for eating soup).

We're going to focus on R because that's a common option in psychological research. Why do people like using R/what makes it well-suited to psychological research? (1) It's open-source and free, meaning users like you and me can add additional functionality on top of base R, and it's free for anyone to use. (2) It has an active teaching community, so a lot of great resources for learning are freely available online. (3) Particularly thanks to Posit/RStudio, it's easy to set up projects and keep files organized. (4) It's excellent at statistical analyses and graphing.

Some other options you may come across are SPSS and JASP, which are "point and click." These are powerful tools and can be totally fine for conducting analyses! They're just a little less flexible because you're limited to their existing functionality. They can also be less reproducible, with people clicking around and not saving their work. There are ways to save your work, which you should do if you're using them.

## Creating Projects

It's important to keep files clean and organized when you're working on a project. I like to create folders dedicated to a given project, and then keep all files and data related to that project in a given folder. Let's create a folder for these workshops. Go to File -\> New Project... -\> New Directory -\> New Project. Create a directory (AKA a folder) and choose where to put it, and click Create Project. This is where your files for these workshops will live! Let's make a file called `intro_to_R.R` (ctrl/cmd-shift-N or File -\> New File -\> R Script) for today's activities.

It's important to keep track of where your files "live" for a given project, as that makes it easier to find your work and have different components interact (e.g., loading a data file in a given script).

Important things in RStudio to start: we have the console (runs code, doesn't save the code anywhere permanent), environment (shows all the objects you've created, right now we have none), and then the files/plots/help window. When you create a file, you also add a script editor.

## Variables, Vectors, and Indexing Vectors

R is a flexible language that can perform a lot of tasks, from basic arithmetic up through advanced statistics analyses and graphing. It is often the case that we want to save the results of some work, either to use later on or just to save us from having to re-run things repeatedly. We can save results to a variable using the assignment operator, \<- . I don't want to get into the nitty-gritty, but \<- is convention when you're assigning values to variables, and = is typically used for function parameter definitions. Most important is that you are consistent.

```{r}
# example 1: creating a variable that contains the result of addition
my_addition <- 35 + 24
print(my_addition)

# example 2: creating a variable that stores a string (text)
favourite_fruit <- "peach"
print(favourite_fruit)
```

Let's take a moment here to talk about some coding best practices: commenting and spacing.

I'm using comments in two ways right now: inline (next to code), and to create sections. Inline, I'm commenting with what I'm doing. This is not just to make it easier for others to read, it's for myself! You know what you're doing when you write your code, but when you come back days, weeks, months, years later, you won't always remember, so it's useful to have a quick description of what is happening and why. As an additional benefit, it can help people reading your code for the first time as well. With creating sections, I'm organizing my file so it's easy to navigate with a legend. These lines with a preceding \# are comments, so won't be run when I run my code. If you follow up a comment with at least four hyphens (# ----), you create a new section. I have a ton of hyphens following my comment because it makes it easier to pick out sections when I scroll through the document, and it's the default of the automatic section creation function in R, which is ctrl-shift-R on a Mac.

Notice that I'm spacing out everything. I have a space around the assignment operator `<-`, spaces between the addition/subtraction/multiplication/division sign and the numbers. This is because it's easier to read. For example, if I write:

```{r}
2+2-1*8/3
```

It is harder to read than:

```{r}
2 + 2 - 1 * 8 / 3
```

This is less noticeable with smaller lines like this, but when you have a bunch of text all at once, it's very noticeable when it's hard to read.

R can store more than single values at once! It can also store multiple elements together, for example the grades in a class or groceries:

```{r}
grades <- c(88, 89, 85, 91)
print(grades)

groceries <- c("cereal", "milk", "bananas", "dragonfruit", "chicken", "basil")
print(groceries)
```

This is very useful for thinking about your data, which is basically a bunch of column vectors (i.e., vectors that are vertically arranged) pushed together. You can index the elements of a vector to see what is stored. In R, indexing starts at 1 (unlike, e.g., Python where the first position is indexed to 0). So to access the value "bananas" from the groceries vector:

```{r}
groceries[3]
```

A vector can only be one data type (integers, doubles \[decimal numbers\], strings). If you try to put multiple types into one list, the data will be coerced into a common type. For example, grocery items and how many of each to get:

```{r}
groceries_numbered <- c("cereal", 1, "milk", 1, "bananas", 7, "dragonfruit", 2, "chicken", 1, "basil", 20)
print(groceries_numbered) # all entries coerced to strings
```

Lists, on the other hand (not vectors), can contain multiple types:

```{r}
groceries_list <- list("cereal", 1, "milk", 1, "bananas", 7)
groceries_list
```

## Functions

Effective use of a programming language involves using the power of functions. We have already seen some functions, like `c()` (combine or concatenate vectors), `print()` (print something to the console), `list()` (create a list). Functions involve three basic ingredients:

1.  Take some input
2.  Do some action on the input
3.  Return an output that results from the action on the input

As a metaphor, think of the function of an oven.

1.  Take some input: raw food
2.  Do some action: get really hot
3.  Return some output: cooked food

When you use an oven, you always put in something you want heated, the oven gets hot (to different degrees, which you can specify depending on what you're putting in it), and it returns the input, but hotter! Functions in programming work like this, too. They take some input, do some action, return some output. For example, the `sum()` function:

1.  Take some input: a vector of numbers
2.  Do some action: add them together
3.  Return some output: the sum of the numbers

You can write your own functions as follows:

```{r}
function_name <- function(variables_for_function) {
  
  # action
  # return
  
}

# here is a fake oven function:
code_oven <- function(food1, food2) {
  
  print("Cooking the foods!")
  finished_food <- c(food1, "and", food2)
  
  return(finished_food)
  
}

code_oven("chicken", "rice")
```

## Installing and Loading Packages

You can use existing functions in base R or create your own functions, and you can *also* use functions written by other people. Recall that R is open-source, meaning anyone can write functions that can be freely used by other people. It's like Wikipedia in that anyone can write something, but there are editors who check that edits make sense. When people publish their functions to be used by others, those functions are contained in *packages*. To use a package and the functions therein, you need to (1) install the package, (2) attach the package to your R session (what we have open right now is an R session).

You only need to install the package once (or when it is updated), but you need to attach it to every session you want to use it in. It's like any other software: you need to download Google Chrome once to have it on your computer (and sometimes you need to redownload to update it), but every time you want to use it you have to open it.

Note that I usually put ALL packages I'll be using at the very top of the document, along with the data I'll be using.

```{r}
# to install, need to do this once
# install.packages("dplyr") # you need the quotation marks
# install.packages("sjPlot")

# to attach to your session, need to do this every session
library(dplyr)
library("dplyr") # you can use quotation marks or not when attaching

library(sjPlot)
```

`dplyr` stands for "data pliers," and is used for manipulating data. Let's look at some example datasets included in the dplyr package, starting with mtcars. This has information about cars' fuel consumption. We will use sjPlot later, when we are presenting the results of our statistical analyses.

## Inspecting Data

It is usually a good idea to look over your data when you read it in. Reasons include (1) familiarizing yourself with its structure, (2) briefly checking it read in correctly.

The mtcars dataset is included in dplyr, which we've loaded, so it's already available to us. A few useful functions for looking at your data in R:

```{r}

head(mtcars) # look at the first few rows
tail(mtcars) # look at the last few rows
summary(mtcars) # get an overview of values
str(mtcars) # overview of data types
View(mtcars) # view the entire dataset

```

## Data Manipulation with `dplyr`

Unlike in Excel, you can't access the data directly when Viewing it. Why? Because it's better to change it using your code in a reproducible way (no wondering if values were changed, forgetting what was changed or moved, etc.), and that's where dplyr comes in! It's an R package used for data manipulation, with a few main functions: filter (choosing certain rows), select (choosing certain columns), mutate (adding/changing columns), arrange (order of rows), summarize (summarizing, as the name suggests).

### Filter

Our dataset has information about cars' number of cylinders. Generally, more cylinders = more power. We can use equality/relational statements to filter the dataset to cars that have more than/less than/equal to a certain number of cylinders.

```{r}

filter(mtcars, cyl == 6) # equal to, notice the two ==
filter(mtcars, cyl < 6) # less than 6
filter(mtcars, cyl > 6) # greater than 6
filter(mtcars, cyl <= 6) # less than or equal to

```

Take cyl == 6. We can see that there are 7 rows. But if we call mtcars, it's still the full dataset.

```{r}

mtcars

```

That's because we didn't save it to a new dataset. If we want to work with a new dataset, we would need to save it to a variable.

```{r}

mtcars_6cyl <- filter(mtcars, cyl == 6)

```

### Select

So filter chooses certain rows. Select chooses certain columns! Let's say we just wanted to look at the car's miles per gallon (mpg) and cylinders (cyl).

```{r}

# either way works
select(mtcars, mpg, cyl)
select(mtcars, c(mpg, cyl))

# again, if we wanted a new dataset, we would have to save it
mtcars_mpg_cyl <- select(mtcars, c(mpg, cyl))

# You can also *deselect* certain columns using a minus sign.
select(mtcars, -mpg) # removes mpg
select(mtcars, -c(mpg, cyl)) # removes mpg and cyl

```

### Mutate

Mutate adds or changes column values. For example, we have the variable qsec which is the time it takes a car to go a quarter of a mile, in seconds. Maybe we want to know how quickly it can go one mile. Let's multiply by four.

```{r}

mutate(mtcars, milesec = 4*qsec)

```

### Group By

Grouping is a powerful operation for data manipulation. Sometimes you want information about an entire group, for example maybe we want the average horsepower across cylinders. To do that, we first group the dataset by number of cylinders, then use the mutate function to calculate average horsepower for each group, and then ungroup and go on with our day.

```{r}

ungroup(mutate(group_by(mtcars, cyl), hp_average = mean(hp)))

```

You might notice that it's getting a little difficult to read! This is where the "pipe" operator comes in. In older versions of R, it looks like this `%>%` and you have to attach the packages dplyr or magrittr to your session to use it. In R 4.1 and newer, there is a pipe built into base R, which looks like `|>`. The pipe makes it easier to read your code, because it takes the result of what you just ran, and passes it on to the next thing you're going to run. Our same operation looks like this using the pipe:

```{r}

mtcars |>                           # take the dataset and pass it forward
  group_by(cyl) |>                  # group the dataset by cyl and pass it forward
  mutate(hp_average = mean(hp)) |>  # calculate the mean horsepower
  ungroup()                          # ungroup the dataset and print the result

# again, if you want to save the result of your calculation, you need to save it
# to a variable

mtcars_hp <- mtcars |>
  group_by(cyl) |>
  mutate(hp_average = mean(hp)) |>
  ungroup()

```

### Arrange

Arrange organizes the rows in ascending or descending order. The default is ascending, i.e., lowest values at the top. If we want to see the lowest horsepower cars first:

```{r}

arrange(mtcars, hp)
mtcars |> arrange(hp) # with the pipe

# highest horsepower, use desc() for descending
arrange(mtcars, desc(hp))
mtcars |> arrange(desc(hp))

```

### Summarize/Summarise (either spelling works)

Summarize creates a new data frame summarizing some data. If the data are ungrouped, it returns one row. If the data are grouped, it returns one row per group. See some functions in the documentation: `?summarize`.

```{r}

mtcars |>
  summarize(mean = mean(hp),
            n = n())

```

There are a bunch of other functions, as well, like renaming columns, removing them, doing rowwise operations, etc. https://dplyr.tidyverse.org/

## Statistical Analyses

Let's do some common statistical analyses using a second dataset, starwars, which is included with dplyr and has information about 14 Star Wars characters like their height, mass, gender, species, etc. I saved it as a csv so we can practice loading files from outside of R.

Option 1: relative filepath, which points to the file *relative* to your current location.

```{r}
starwars <- read.csv("starwars.csv")
```

Your current directory is the starting point for where your computer will look for the file. If `starwars.csv` is in the current folder, then great, your computer will find it! If you created a sub-folder, e.g., `data`, then you need to specify that your computer needs to search there for the file: `read.csv("data/starwars.csv")`.

Option 2: absolute filepath. NOT RECOMMENDED. Your code will not work when someone else loads it because it will be referencing a specific filepath on *your* computer.

```{r}
# starwars <- read.csv("/Users/yourname/R_tutorial/starwars.csv")
```

Let's inspect our data to see if it read in correctly:

```{r}

head(starwars) # first few rows
summary(starwars) # summaries of variables
str(starwars) # structures

```

We'll focus on a few analyses that you might encounter or that you've seen before: t-tests, ANOVAs, linear and logistic regression. Then we'll look at displaying results nicely with `sjPlot`. Note that this code covers how to conduct a variety of analyses quite quickly; the theory underlying these analyses is beyond the scope of this work.

### t-test

Let's use a t-test to assess whether humans' weights are significantly different from other species.

```{r}

starwars <- starwars |> 
  mutate(human = ifelse(species == "Human", "human", "nonhuman")) |> 
  mutate(human_droid_other = case_when(species == "Human" ~ "human",
                                       species == "Droid" ~ "droid",
                                       .default = "other"))

```

In general, R formulas are "Y \~ X", which you can read as something like "Y predicted by X". So here, "mass 'predicted by' human status," which is to say "how does mass differ between the human and non-human species?"

```{r}

t.test(mass ~ human, data = starwars)

```

### Linear regression

You can conduct a linear regression using the `lm()` function, also in the base `stats` package. Let's look at how height and mass are related.

```{r}
# simple linear regression
model1 <- lm(mass ~ height, data = starwars) 
summary(model1)

# We can add other predictors for multiple linear regression
model2 <- lm(mass ~ height + gender, data = starwars) 
summary(model2)

```

### ANOVA

A linear regression with a categorical predictor is an anova, but for completeness we can look here at conducting an ANOVA and doing some post-hoc tests. For example, let's look at the effect of human/droid/other species status on mass.

```{r}
# anova
anova <- aov(mass ~ human_droid_other, data = starwars)
summary(anova)

# post-hoc tests (our result isn't significant, but just to illustrate how to do it)
TukeyHSD(anova)

# also note that a linear regression with a categorical predictor is just an anova, simpler to use one framework
anova_lm <- lm(mass ~ human_droid_other, data = starwars)
summary(anova_lm)
```

We won't discuss assumptions, but you can check with this package (https://easystats.github.io/performance/articles/check_model.html) or with any number of other options in R.

You can print your model results in nice tables and visualize your regression coefficients using the `sjPlot` package. In particular, the function `tab_model()` prints a table ("tab") for your model, and `plot_model()` generates a plot.

```{r}
# Presenting Results: sjPlot and ggplot2 ----------------------------------

# install.packages(c("sjPlot", "ggplot2")) # you have to do this once
library(sjPlot) # have to attach to session every time

# For our simple linear regression, model
tab_model(model1)
plot_model(model1)

# For our multiple linear regression, model2
tab_model(model2)
plot_model(model2)

# For our simple linear regression, categorical predictor, anova_lm
tab_model(anova_lm)
plot_model(anova_lm)

# For our anova
tab_model(anova)
```

## If time, graphing with ggplot2

ggplot stands for "grammar of graphics" plotting, which is the idea that just like languages have grammars and once you know that you can construct totally new sentences, graphs also have a "grammar," fundamental elements that you can use to build all kinds of graphs.

The main ingredients in a graph are (1) the plotting region (this would be like a piece of paper, in the physical world) and (2) the shapes you're using in your graph (this is your type of graph) In ggplot2, the shapes are generally called "geom"s, and some examples of geoms are histograms, scatterplots, bar graphs, line graphs, boxplots, violin plots.

```{r}
library(ggplot2) # again, I would usually load all packages at the top of a script

ggplot(starwars) + # this says "I'm making a plot", like getting out a piece of paper
  geom_point(mapping = aes(x = height, y = mass)) # this is the "point" geom, for a scatterplot. mapping = aes(...) describes what's on your axes
```

You can use `mapping = aes()` in the `ggplot()` function, in which case it applies to ALL geoms unless overwritten, or in an individual geom (just applies to that geom).

There's a lot of graphing, we're not going to come close to doing everything possible. There are some resources in the tutorial notes document for further learning. The main things we're going to touch on are a few basic shapes and colours.

```{r}
# Graph of height by gender.
ggplot(starwars, mapping = aes(x = gender, y = height)) +
  geom_col()

# Some geoms distinguish between colour (the outside edge) and fill (the inside)

ggplot(starwars, mapping = aes(x = gender, colour = gender)) +
  geom_bar()

ggplot(starwars, mapping = aes(x = gender, fill = gender)) +
  geom_bar()

# some don't distinguish
ggplot(starwars) +
  geom_point(mapping = aes(x = height, y = mass, colour = gender))

```

You can also do data manipulation and then pipe into a graph. NOTE THE DIFFERENCE: ggplot uses plus signs (+) to join things together, not pipes (%\>% or \|\>).

```{r}
starwars |>  
  ggplot() +
  geom_point(mapping = aes(x = height, y = mass, colour = gender))

starwars |>
  filter(!is.na(gender)) |>
  ggplot() +
  geom_point(mapping = aes(x = height, y = mass, colour = gender))
```
